Approximation as Differentiation

### 2025-05-24 main/formal-structures.tex
Say something about CC on Fam(X).
Start on todo's for new section.

In these special cases when $\cat{C} = \Set$, the Grothendieck construction is the {\em families} constructions $\Fam(\FinVect)$ and $\Fam(\LatGal)$~(\secref{Fam}). These settings are adequate for interpreting first-order programs, but neither is Cartesian closed. To interpret higher-order programs in $\Fam(\cat{X})$ for any category $\cat{X}$, richer structure on $\cat{X}$ is required. As is well known~\cite{lawvere63}, $\Fam(\cat{X})$ is the free coproduct completion of $\cat{X}$. In the case that $\cat{X}$ has a (symmetric) monoidal product, then we always get a symmetric monoidal product on $\Fam(\cat{X})$ (using the Cartesian products from $\Set$). When $\cat{X}$ has all (small) products, and the monoidal product is actually a coproduct, then $\Fam(\cat{X})$ is symmetric monoidal closed. If the coproducts are actually {\em biproducts}~(\secref{biproducts}), then $\Fam(\cat{X})$ is Cartesian closed, and we can interpret higher-order programs.

Happily, $\FinVect$ and $\LatGal$ both have biproducts, as a consequence of their having products and being enriched in commutative monoids~(\propref{biproducts:from-product-or-coproduct} below). However, neither of them has all small products, without moving to infinite dimensional vector spaces (in the case of $\FinVect$) or complete lattices (in the case of $\LatGal$). Neither of these are ideal. In infinite dimensional vector spaces, dualisation is not involutive and the connection between the forward and reverse derivatives is lost. In complete lattices, we can no longer easily implement the infinite meets and joins required, and we lose the computability of the model. Moreover, in Agda, there are issues with predicativity when considering complete lattices.\todo{be more specific}

Vákár overcomes this by using different interpretations for forward and backward derivatives. we overcome this by using $\CMon$-enriched presheaves.\todo{elaborate.}

@@

\subsection{Biproducts and Semi-Additive Categories}
\label{sec:biproducts}

We noted that $\LatGal$ is enriched in $\SemiLat$, the category of semilattices, and moreover that the product
of lattices is also a coproduct, making $\LatGal$ into a category with \emph{biproducts}. In fact the presence
of biproducts is sufficient for enrichment in $\CMon$, the category of commutative monoids (including
semilattices); such a category is called \emph{semi-additive}.

\begin{definition}[Semi-additive category]
\label{def:biproducts:semi-additive}
A category with finite products and coproducts is \emph{semi-additive} if the canonical morphisms (projections
and injections) give an isomorphism
\[\textstyle X \coprod Y \iso X \times Y\] that is natural in both variables.
\end{definition}

The product/coproduct is called a \emph{biproduct}, with the biproduct structure denoted by $(\biprod, 0)$ and
projections $\biproj_X, \biproj_Y$ and injections $\biinj_X$ and $\biinj_Y$. The unit $0$ is both terminal and
initial and is called a \emph{zero} object. A semi-additive category $\cat{C}$ is enriched in $\CMon$: for any
two morphisms $f, g: X \to Y$ in $\cat{C}$, the biproduct structure provides a way to ``add'' them together,
forming a morphism $f + g: X \to Y$. Diagrammatically:

\begin{center}
\begin{tikzcd}
   X \arrow[r, "\diag"] & X \biprod X \arrow[r, "f \biprod g"] & Y \biprod Y \arrow[r, "\codiag"] & Y
\end{tikzcd}
\end{center}

Here $\diag$ denotes the diagonal $\prodM{\id_X}{\id_X}$ given by the universal property of the product and
$\codiag$ denotes the codiagonal $\coprodM{\id_X}{\id_Y}$ given by the universal property of the coproduct.
The $f \oplus g$ morphism is the component-wise map $\prodM{f \comp \biproj_X}{g \comp \biproj_Y} =
\coprodM{\biinj_X \comp f}{\biinj_Y \comp g}$. Similarly we can exhibit a \emph{zero} morphism $0_{X,Y}$ by
composing the unique maps in and out of the zero object:

\begin{center}
\begin{tikzcd}
   X \arrow[r, "!_X"] & 0 \arrow[r, "!^Y"] & Y
\end{tikzcd}
\end{center}

It is easy to verify that $+$ is associative and commutative and that $f + 0_{X,Y} = f$, and thus that every
hom-object $\cat{C}(X,Y)$ is an object in $\CMon$. Moreover composition is \emph{bilinear}, i.e.~given by a
family of morphisms $\Hom{\cat{C}}{Y}{Z} \tensor \Hom{\cat{C}}{X}{Y} \to \Hom{\cat{C}}{X}{Z}$ in $\CMon$ that
preserve the additive structure in $\Hom{\cat{C}}{Y}{Z}$ and $\Hom{\cat{C}}{X}{Y}$ separately:

\begin{salign*}
f \comp \zero_{X,Y} = 0_{X,Z} = \zero_{Y,Z} \comp f
\end{salign*}
\begin{salign*}
(f + g) \comp h &= (f \comp h) + (g \comp h) \\
h \comp (f + g) &= (h \comp f) + (h \comp g)
\end{salign*}

\subsubsection{Biproduct laws}
It is also easy to show that in a semi-additive category, as defined in \defref{biproducts:semi-additive}, the
following equations hold:

\vspace{-3mm}
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_X \comp \biinj_X &= \id_X \\
   \biproj_Y \comp \biinj_X &= \zero_{X,Y}
\end{salign*}
\end{center}
\end{minipage}%
\begin{minipage}[t]{0.45\textwidth}
\begin{center}
\begin{salign*}
   \biproj_Y \comp \biinj_Y &= \id_Y \\
   \biproj_X \comp \biinj_Y &= \zero_{Y,X}
\end{salign*}
\end{center}
\end{minipage}

\noindent and

\begin{salign*}
(\biinj_X \comp \biproj_X) + (\biinj_Y \comp \biproj_Y) &= \id_{X \biprod Y}
\end{salign*}

We can alternatively take these to be the defining equations of a biproduct $(X \biprod Y, \biproj_X,
\biproj_Y, \biinj_X, \biinj_Y)$ in any category $\cat{C}$ known to be $\CMon$-enriched, and therefore
providing addition of morphisms; in that case any products or coproducts in $\cat{C}$ are necessarily
biproducts.

\begin{proposition}
\label{prop:biproducts:from-product-or-coproduct}
Suppose $\cat{C}$ is $\CMon$-enriched.
\begin{enumerate}
\item If $(X \times Y, \pi_1, \pi_2)$ is a product in $\cat{C}$ then $(X \times Y, \biinj_X, \biinj_Y, \proj_1, \proj_2)$
is a biproduct, where $\biinj_X = \prodM{\id_X}{0_{X,Y}}: X \to X \times Y$ and $\biinj_Y =
\prodM{0_{X,Y}}{\id_Y}: Y \to X \times Y$.
\item If $(X \coprod Y, \inj_1, \inj_2)$ is a coproduct in $\cat{C}$ then $(X \coprod Y, \inj_1, \inj_2,
\biproj_X, \biproj_Y)$ is a biproduct, where $\biproj_X = \coprodM{\id_X}{0_{Y,X}}: X \coprod Y \to X$ and
$\biproj_Y = \coprodM{0_{X,Y}}{\id_Y}: X \coprod Y \to Y$.
\item If $X$ is terminal (or initial) in $\cat{C}$ then it is a zero object.
\end{enumerate}
\end{proposition}

>>>>

In these special cases when $\cat{C} = \Set$, the Grothendieck construction is the {\em families} constructions $\Fam(\FinVect)$ and $\Fam(\LatGal)$~(\secref{Fam}). These settings are adequate for interpreting first-order programs, but neither is Cartesian closed. To interpret higher-order programs, the picture is more complicated. As is well known, $\Fam(\cat{X})$ for any category $\cat{X}$ is the free coproduct completion of $\cat{X}$~\cite{lawvere63}. In the case that $\cat{X}$ has a (symmetric) monoidal product, then we always get a symmetric monoidal product on $\Fam(\cat{X})$ (using the Cartesian products from $\Set$). When $\cat{X}$ has all (small) products, and the monoidal product is actually a coproduct, then $\Fam(\cat{X})$ is symmetric monoidal closed. If the coproducts are actually {\em biproducts}~(\secref{biproducts}), then $\Fam(\cat{X})$ is Cartesian closed, and we can interpret higher-order programs.

Happily, $\FinVect$ and $\LatGal$ both have biproducts, as a consequence of their having products and being enriched in commutative monoids~(\propref{biproducts:from-product-or-coproduct} below). However, neither of them has all small products, without moving to infinite dimensional vector spaces (in the case of $\FinVect$) or complete lattices (in the case of $\LatGal$). Neither of these are ideal. In infinite dimensional vector spaces, dualisation is not involutive and the connection between the forward and reverse derivatives is lost. In complete lattices, we can no longer easily implement the infinite meets and joins required, and we lose the computability of the model. Moreover, in Agda, there are issues with predicativity when considering complete lattices. \todo{be more specific}

Vákár and collaborators overcome this by using different interpretations for forward and backward derivatives. We overcome this by using $\CMon$-enriched presheaves; we turn to this in \secref{}, but first we flesh out the first-order setting.

@@
